---
title: "Visium_spotlight"
output: html_document
---

```{=html}
<style type="text/css">
.smaller {
  font-size: 10px
}
</style>
```

```{r, setup,  warning = FALSE}
setwd("/data/home/lyang/Visium_spotlight")
  knitr::opts_knit$set(root.dir = "/data/home/lyang/Visium_spotlight")


```


```{r}
# location = "server"
# 
# if(location == "my_computer"){
#   setwd("C:/Users/12895/Documents")
#   resultsdir <- 'cell2location_Visium_RCTD' 
#   dir.create(resultsdir)
#   knitr::opts_knit$set(root.dir = 'C:/Users/12895/Documents/cell2location_Visium_spotlight')
# }else if(location == "server"){
#   setwd("/data/home/lyang")
#   resultsdir <- 'cell2location_Visium_spotlight' 
#   dir.create(resultsdir)
#   knitr::opts_knit$set(root.dir = "/data/home/lyang/cell2location_Visium_spotlight")
# 
# }


```

# Load packages {.unnumbered}

```{r load-libs, message = FALSE,  warning = FALSE}

#! /usr/bin/Rscript
# screen -S spotlight_cell2location34
# Rscript run_spotlight.R &>nohup.out&
  
# .libPaths()
# 
# 
# 
# if (!require("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# 
# BiocManager::install("SpatialExperiment")


.libPaths(c("/data/home/lyang/R/x86_64-redhat-linux-gnu-library/4.1",
            "/data/home/bioinfo/R/R4.1.0/library_B3.13",
            "/usr/lib64/R/library","/usr/share/R/library") )
dyn.load("/data/home/bioinfo/programs/hdf5-1.12.0/hdf5/lib/libhdf5_hl.so.200")


init_require_packages <- function(){
  library(ggplot2)
  library(SPOTlight)
  library(SingleCellExperiment)
  library(SpatialExperiment)
  library(scater)
  library(scran)
  library(Seurat)
  library(SeuratData)
  library(SeuratDisk)
}

init_require_packages()


```


```{r eh, message = FALSE}
# spatial_file = "Visium_spatial"
# paste(spatial_file,"rds",sep =".")
```

Load the spatial data:

```{r load-sp, message=FALSE,results = 'hide', warning = FALSE}


if(file.exists("Visium_spatial.rds")){
  rds_file='Visium_spatial.rds'
  spatial <- readRDS(rds_file)
}else{
  data.dir = 'D:/minor_intership_data/Visium_Spatial_Lymph_Node_Data/data'
  
  spatial = Load10X_Spatial( data.dir , filename =
                               "V1_Human_Lymph_Node_filtered_feature_bc_matrix.h5" )
  saveRDS(spatial,"Visium_spatial.rds")
}
```

Load the single cell data.

```{r load-sc, message=FALSE}
# dataset_name = "cell2location" Tabula
dataset_name = "cell2location34"
sc_dataset_file = paste(dataset_name,"sc_dataset.rds",sep ="_")

# library(SeuratDisk)
if(file.exists(sc_dataset_file)){
  rds_file= sc_dataset_file
  sc_dataset <- readRDS(rds_file)
}else{
  import_sc_dataset <- function()
  {
    # Convert("D:/minor_intership_data/TS_Lymph_Node.h5ad", dest = "h5seurat", overwrite = TRUE)
    sc_dataset <- LoadH5Seurat("D:/minor_intership_data/TS_Lymph_Node.h5seurat",assays  = "RNA")
    sc_dataset
    head(sc_dataset@meta.data, 5)
    saveRDS(sc_dataset,'sc_dataset.rds')
    return(sc_dataset)
  }
  sc_dataset = import_sc_dataset()
}  

# colnames(sc_dataset@meta.data) 
# table(sc_dataset@meta.data$CellType2)
# table(sc_dataset@meta.data$Subset)


```

Quick data exploration:

```{r explo, fig.width=12, fig.height=6}
# annotation = "cell2location34" Tabula
# annotation = "cell2location34"
dataset_name = "cell2location34"
if(dataset_name == "cell2location34"){
  
  cell_type_table = as.data.frame(table(sc_dataset$Subset))
    
}else if(dataset_name == "Tabula"){
  cell_type_table = as.data.frame(table(sc_dataset$cell_ontology_class))
}else if(dataset_name == "cell2location44"){
  cell_type_table = as.data.frame(table(sc_dataset$CellType))
}
  



colnames(cell_type_table) = c("cell_type","frequency")
library(ggplot2)
# Basic barplot
p <- ggplot(data=cell_type_table, aes(x=cell_type, y=frequency,fill=cell_type)) +
  geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))+
       geom_text(aes(label=frequency), position=position_dodge(width=0.9), vjust=-0.25)


ggsave(p,filename = "cell_type_histogram.pdf",width = 15,
       height = 20)

print(p)

```



```{r subset}

# Keep cells with clear cell type annotations
# sce <- subset(sce, , !cell_ontology_class %in% c("nan", "CD45"))
# sc_dataset <- subset(sc_dataset, subset = cell_ontology_class != "nan")

 

```



```{r lognorm}

sc_reference_markers_file = paste(dataset_name,"sc_reference_markers.rds",sep ="_")
sc_hvg_file = paste(dataset_name,"sc_hvg.rds",sep ="_")

if(file.exists(sc_hvg_file)){
  rds_file= sc_reference_markers_file
  sc_reference_markers <- readRDS(rds_file)
  rds_file= sc_hvg_file
  sc_hvg <- readRDS(rds_file)
}else
  {
  
  qc_filter_sc <- function(sc_dataset,dataset_name){
    if(dataset_name == "Tabula"){
      sc_dataset[['nCount_RNA']] = sc_dataset$n_counts_UMIs
    }else if(dataset_name == "cell2location"){
      sc_dataset[['nCount_RNA']] = sc_dataset$n_counts
    }
    # head(Tabula_sc@meta.data)
    # The [[ operator can add columns to object metadata. This is a great place to stash QC stats
    # Tabula_sc[["percent.mt"]] <- PercentageFeatureSet(Tabula_sc, pattern = "^MT-")
    mt.genes <- grep(pattern = '^MT-', x = rownames(x = sc_dataset@assays$RNA),value = TRUE)
    
    percent.mt <- Matrix::colSums(sc_dataset@assays$RNA[mt.genes, ]) / Matrix::colSums(sc_dataset@assays$RNA)
    
    sc_dataset <- AddMetaData(object = sc_dataset,metadata = percent.mt,
                              col.name = 'percent.mt')
    
    
    # filter cells based on QC metrics
    sc_dataset_filtered <- subset(sc_dataset, subset = n_genes > 200 & n_genes < 5000 & percent.mt < 0.05 )
    return(list("sc"= sc_dataset,"sc_filtered"= sc_dataset_filtered))
    
  }
  
  if(dataset_name=="cell2location" | dataset_name == "cell2location34"){
    a_list = qc_filter_sc(sc_dataset,"cell2location")
  }else if(dataset_name=="Tabula"){
    a_list = qc_filter_sc(sc_dataset,"Tabula")
  }
  
  sc_dataset = a_list$sc
  # Visualize QC metrics as a violin plot
  VlnPlot(sc_dataset, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
  sc_dataset_filtered = a_list$sc_filtered
  sc_dataset_filtered <- NormalizeData(sc_dataset_filtered, normalization.method = "LogNormalize", scale.factor = 10000)
  head(rownames(sc_dataset_filtered))
  # Get vector indicating which genes are neither ribosomal or mitochondrial
  genes <- !grepl(pattern = "^Rp[l|s]|Mt", x = rownames(sc_dataset_filtered))
  if(dataset_name=="Tabula"){
    Idents(sc_dataset_filtered) <- 'cell_ontology_class'
  }else if(dataset_name=="cell2location"){
    Idents(sc_dataset_filtered) <- 'CellType'
  }else if(dataset_name=="cell2location34"){
    Idents(sc_dataset_filtered) <- 'Subset'
  }
  
  head(Idents(sc_dataset_filtered), 5)
  sc_dataset_filtered <- FindVariableFeatures(sc_dataset_filtered, selection.method = "vst", nfeatures = 3000)
  sc_hvg =VariableFeatures(sc_dataset_filtered)
  saveRDS(sc_hvg,sc_hvg_file)
  sc_reference_markers <- FindAllMarkers(sc_dataset_filtered, min.pct = 0.25,
                                         logfc.threshold = 0.25,max.cells.per.ident=200)
  library(dplyr)
  sc_reference_markers  <- sc_reference_markers %>%
    group_by(cluster) %>%
    slice_max(n = 10, order_by = avg_log2FC)
  saveRDS(sc_reference_markers,sc_reference_markers_file)
  
  
}




```

### Cell Downsampling
```{r downsample}

sc_dataset.downsampled_file = paste(dataset_name,"sc_dataset.downsampled.rds",sep ="_")

if (file.exists(sc_dataset.downsampled_file)){
  rds_file = sc_dataset.downsampled_file
  sc_dataset_filtered <- readRDS(rds_file)
}else
{
  # downsample to at most 100 per identity
  subsample_cells <- function(sc_dataset,dataset_name)
  {
    
    if(dataset_name == "Tabula"){
      Idents(sc_dataset) <- 'cell_ontology_class'
    }else if(dataset_name == "cell2location"){
      Idents(sc_dataset) <- 'CellType'
    }else if(dataset_name == "cell2location34"){
      Idents(sc_dataset) <- 'Subset'
    }
    table(Idents(sc_dataset))
    unique(Idents(sc_dataset))
    
    
    cell.list <- WhichCells(sc_dataset, idents = unique(Idents(sc_dataset)), 
                            downsample = 1000)
    gc()
    sc_dataset <- sc_dataset[, cell.list]
    # table(sc_dataset$cell_ontology_class)
    saveRDS(sc_dataset,sc_dataset.downsampled_file)
    
    return(sc_dataset)
    
  }
  # subsample fixed numbers of cells from differently sized cell types in a seurat object.
  sc_dataset_filtered = subsample_cells(sc_dataset_filtered,dataset_name)
}


```

## Deconvolution

You are now set to run `SPOTlight` to deconvolute the spots!




```{r SPOTlight}

res_file = paste(dataset_name,"res.rds",sep ="_")

if(file.exists(res_file)){
  rds_file = res_file
  res <- readRDS(rds_file)
}else{
  if(dataset_name=="cell2location34"){
    res <- SPOTlight(
      x = sc_dataset_filtered,
      y = spatial@assays$Spatial@counts,
      groups = sc_dataset_filtered$Subset,
      mgs = sc_reference_markers,
      hvg = sc_hvg,
      weight_id = "avg_log2FC",
      group_id = "cluster",
      gene_id = "gene")
    saveRDS(res,res_file)
  }else if(dataset_name=="Tabula")
  {
    res <- SPOTlight(
      x = sc_dataset_filtered,
      y = spatial@assays$Spatial@counts,
      groups = sc_dataset_filtered$cell_ontology_class,
      mgs = sc_reference_markers,
      hvg = sc_hvg,
      weight_id = "avg_log2FC",
      group_id = "cluster",
      gene_id = "gene")
    saveRDS(res,res_file)
  }
  
  
}
# Time for training: 708.35min
# Deconvoluting mixture data


```

Extract data from `SPOTlight`:

```{r}
# Extract deconvolution matrix
head(mat <- res$mat)[, seq_len(6)]
# Extract NMF model fit
mod <- res$NMF




library(NMF)
sign <- basis(mod)
colnames(sign) <- paste0("Topic", seq_len(ncol(sign)))
head(sign)



```

# Visualization

In the next section we show how to visualize the data and interpret
`SPOTlight`'s results. 

## Topic profiles

We first take a look at the Topic profiles. By setting `facet = FALSE` we want to
evaluate how specific each topic signature is for each cell identity.
Ideally each cell identity will have a unique topic profile associated to it as 
seen below.
```{r , fig.width=12, fig.height=7}
      

```


```{r plotTopicProfiles1}
if(dataset_name == "Tabula"){
  plotTopicProfiles(
    x = mod,
    y = sc_dataset_filtered$cell_ontology_class,
    facet = FALSE,
    min_prop = 0.01,
    ncol = 1) +
    theme(aspect.ratio = 1,axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))
  
  
}else if(dataset_name == "cell2location"){
  plotTopicProfiles(
    x = mod,
    y = sc_dataset_filtered$CellType,
    facet = FALSE,
    min_prop = 0.01,
    ncol = 1) +
    theme(aspect.ratio = 1,axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))
}else if(dataset_name == "cell2location34"){
  plotTopicProfiles(
    x = mod,
    y = sc_dataset_filtered$Subset,
    facet = FALSE,
    min_prop = 0.01,
    ncol = 1) +
    theme(aspect.ratio = 1,axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))
  
}


```

Next we also want to ensure that all the cells from the same cell identity share 
a similar topic profile since this will mean that `SPOTlight` has learned a
consistent signature for all the cells from the same cell identity.

```{r plotTopicProfiles2, fig.width=20, fig.height=30}
if(dataset_name == "Tabula"){
  cell_type = sc_dataset_filtered$cell_ontology_class
}else if(dataset_name == "cell2location"){
  cell_type = sc_dataset_filtered$CellType
}else if(dataset_name == "cell2location34"){
  cell_type = sc_dataset_filtered$Subset
}

plotTopicProfiles(
  x = mod,
  y = cell_type,
  facet = TRUE,
  min_prop = 0.01,
  ncol = 5)

```

Lastly we can take a look at which genes the model learned for each topic.
Higher values indicate that the gene is more relevant for that topic. 
In the below table we can see how the top genes for `Topic1` are characteristic
for B cells (i.e. *Cd79a*, *Cd79b*, *Ms4a1*...).

```{r basis-dt, message=FALSE, warning=FALSE,results = 'hide'}
library(NMF)
sign <- basis(mod)
colnames(sign) <- paste0("Topic", seq_len(ncol(sign)))
head(sign)
# This can be dynamically visualized with DT as shown below
# DT::datatable(sign, fillContainer = TRUE, filter = "top")
```

## Spatial Correlation Matrix

See [here](http://www.sthda.com/english/wiki/ggcorrplot-visualization-of-a-correlation-matrix-using-ggplot2) 
for additional graphical parameters.

```{r plotCorrelationMatrix, fig.width=9, fig.height=9}
# plotCorrelationMatrix(mat)
```

## Co-localization

Now that we know which cell types are found within each spot we can make a graph
representing spatial interactions where cell types will have stronger edges
between them the more often we find them within the same spot.

See [here](https://www.r-graph-gallery.com/network.html) for additional
graphical parameters.

```{r plotInteractions, fig.width=9, fig.height=9}
# plotInteractions(mat, "heatmap")
# plotInteractions(mat, "network")
```


```{r, fig.width=12, fig.height=9}

DimPlot(sc_dataset_filtered, group.by = "Subset", label = TRUE)
# DimPlot(sc_dataset, group.by = "cell_ontology_class")
# PCAPlot(sc_dataset, group.by = "cell_ontology_class")
# UMAPPlot(sc_dataset, group.by = "cell_ontology_class")
```
## Scatterpie

We can also visualize the cell type proportions as sections of a pie chart for 
each spot. You can modify the colors as you would a standard .

```{r Scatterpie, fig.width=12, fig.height=9}
# This can be dynamically visualized with DT as shown below
ct <- colnames(mat)
mat[mat < 0.1] <- 0


library(RColorBrewer)

if(dataset_name == "Tabula"){
  n <- 29
}else if(dataset_name == "cell2location"){
  n <- 44
}else if(dataset_name == "cell2location34"){
  n <- 34
}

  
colrs <- brewer.pal.info[brewer.pal.info$colorblind == TRUE, ]
col_vec = unlist(mapply(brewer.pal, colrs$maxcolors, rownames(colrs)))
set.seed(123)
col <- sample(col_vec, n)




# Define color palette
# (here we use 'paletteMartin' from the 'colorBlindness' package)
paletteMartin <- col

pal <- colorRampPalette(paletteMartin)(length(ct))
names(pal) <- ct

plotSpatialScatterpie(
  x = spatial,
  y = mat,
  cell_types = colnames(y),
  img = FALSE,
  scatterpie_alpha = 1,
  pie_scale = 0.4) +
  scale_fill_manual(
    values = pal,
    breaks = names(pal))

```

## Residuals

Lastly we can also take a look at how well the model predicted the proportions
for each spot. We do this by looking at the residuals of the sum of squares for
each spot which indicates the amount of biological signal not explained by the model.

```{r message=FALSE}
spatial_dir = 'D:/minor_intership_data/Visium_Spatial_Lymph_Node_Data/data'



spatial <- read10xVisium(spatial_dir,
                         type = "HDF5", data = "raw",
                        images = "lowres",load = FALSE)
saveRDS(spatial,"spatial_residual.rds")
spatial$res_ss <- res[[2]][colnames(spatial)]
xy <- spatialCoords(spatial)
spatial$x <- xy[, 1]
spatial$y <- xy[, 2]
ggcells(spatial, aes(x, y, color = res_ss)) +
    geom_point() +
      scale_color_viridis_c() +
    coord_fixed() +
    theme_bw()
```

```{r, warning=FALSE, fig.width=10, fig.height=10}

rds_file='predictions.assay.rds'
predictions.assay <- readRDS(rds_file)


predictions.assay@data <- t(mat)
meta.features <- as.data.frame(colnames(mat) )
rownames(meta.features) = meta.features[,1]
meta.features$`colnames(mat)` =NULL
predictions.assay@meta.features = meta.features


spatial[["predictions"]] <- predictions.assay

DefaultAssay(spatial) <- "predictions"
table(sc_dataset@meta.data$Subset)


if(dataset_name == "Tabula"){
  
p = SpatialFeaturePlot(spatial, features = c("stromal cell", "t cell"), pt.size.factor = 1.6, ncol = 2, crop = TRUE)
print(p)
p = SpatialFeaturePlot(spatial, features = c("naive b cell", "b cell"), pt.size.factor = 1.6, ncol = 2, crop = TRUE)
print(p)
  
}else if(dataset_name == "cell2location"){
  
 SpatialFeaturePlot(spatial, features = c("T_Treg", "DC_cDC2"), pt.size.factor = 1.6, ncol = 2, crop = TRUE)
SpatialFeaturePlot(spatial, features = c("DC_cDC1", "DC_cDC2"), pt.size.factor = 1.6, ncol = 2, crop = TRUE)

}else if(dataset_name == "cell2location34"){
  
 
  SpatialFeaturePlot(spatial, features = c("T_CD4+", "NKT","DC_cDC1"), pt.size.factor = 1.6, ncol = 3, crop = TRUE)
print(p)
p = SpatialFeaturePlot(spatial, features = c("B_GC_prePB", "ILC"), pt.size.factor = 1.6, ncol = 2, crop = TRUE)
print(p)
SpatialFeaturePlot(spatial, features = c("T_CD8+_CD161+", "T_CD8+_cytotoxic"), pt.size.factor = 1.6, ncol = 2, crop = TRUE)
}
# table(sc_dataset@meta.data$Subset)

```

# Session information

```{r session-info}
sessionInfo()
```
